---
title: 'Text analysis workshop: Getting data from PDFs'
author: "Casey O'Hara"
format: 
  html:
    toc: true
    embed-resources: true
    number-sections: true
execute:
  warning: false
  message: false
---

# Overview

How often have you run across a published paper with awesome-looking data, but it is only available in PDF format?  ARGHHH!  But using the `pdftools` package and some `stringr` functions with regex patterns, we can get that data out and into a usable format.

```{r}
library(tidyverse)

library(pdftools)

# ?pdftools

```

# The `pdftools` package

The pdftools package basically has five functions:

* `pdf_info(pdf, opw = "", upw = "")` to get metadata about the pdf itself
* `pdf_text(pdf, opw = "", upw = "")` to get the text out of the pdf.
* `pdf_fonts(pdf, opw = "", upw = "")` to find out what fonts are used (including embedded fonts)
* `pdf_attachments(pdf, opw = "", upw = "")` umm attachments I guess?
* `pdf_toc(pdf, opw = "", upw = "")` and a table of contents.

Really we'll just focus on `pdf_text()`.

``` {r}
pdf_smith <- file.path('pdfs/smith_wilen_2003.pdf')

smith_text <- pdf_text(pdf_smith)
```

`pdf_text()` returns a vector of strings, one for each page of the pdf.  So we can mess with it in tidyverse style, let's turn it into a dataframe, and keep track of the pages.

Then we can use `stringr::str_split()` to break the pages up into individual lines.  Each line of the pdf is concluded with a backslash-n, so split on this.  We will also add a line number in addition to the page number.

``` {r}

smith_df <- data.frame(text = smith_text) # one row per page

smith_df <- data.frame(text = smith_text) %>%
  mutate(page = 1:n()) %>%
  mutate(text_sep = str_split(text, '\\n')) %>% # split by line, text_set=lists of lines 
  unnest(text_sep) # separate lists into rows

smith_df <- data.frame(text = smith_text) %>%
  mutate(page = 1:n()) %>%
  mutate(text_sep = str_split(text, '\\n')) %>%
  unnest(text_sep) %>%
  group_by(page) %>%
  mutate(line = 1:n()) %>% # add line #s by page
  ungroup()

```

# Getting the table out of the PDF

Let's look at the PDF, specifically we want to get the data in the table on page 8 of the document. More specifically, the table data is in lines 8 to 18 on page 8.  This is a table comparing the number of active urchin divers to the number of patches in which they dove for urchins, from 1988 to 1999.

```{r}
smith_df %>%
  filter(page == 8 & line <= 25 & line >= 7) %>%
  pull(text_sep)
```


* The column headings are annoyingly just years, and R doesn't like numbers as column names, so we'll rename them as 'y####' for year.
* Break up the columns separated by (probably tabs but we can just use) spaces.  We'll use the `tidyr::separate` function to separate into columns by spaces.  Note, one space and multiple spaces should count the same way - how to do that in regex?

## extract table (not run in workshop)

``` {r}
### We want to extract data from the table on page 8
page8_df <- smith_df %>%
  filter(page == 8)

### Let's just brute force cut out the table
col_lbls <- c('n_patches', paste0('y', 1988:1999))

table1_df <- page8_df %>%
  filter(line %in% 8:18) %>%
  separate(text_sep, col_lbls, ' +') 

```

Now we can ditch the `text`, `page`, and `line` columns, and pull the result into a tidy format (long format rather than wide) for easier operations.

* When we pull the 'y####' columns into a year column, let's turn those into integers instead of a character
* Same goes for the number of patches and number of divers - they're all character instead of integer (because it started out as text).

``` {r}

table1_tidy_df <- table1_df %>%
  select(-text, -line, -page) %>%
  gather(year, n_divers, starts_with('y')) %>%
  mutate(year = str_replace(year, 'y', ''), ### or str_extract(year, '[0-9]{4}')
         year = as.integer(year),
         n_patches = as.integer(n_patches),
         n_divers = as.integer(n_divers))

DT::datatable(table1_tidy_df)
```

With pdftools, we extracted a data table, but we could also just extract the text itself if that's what we really wanted.  

## See also: https://github.com/ropensci/tabulizer
